Entrainment of low Mach number thermals in stratified domains

### 2019-05-29 tex/stratified_thermals.tex
cleans up manuscript per suggestions from Daniel + cleans up code some

\begin{abstract}
% first sentence should state the solar convective conundrum. Something like: Historically, the sun's luminosity was thought to be carried by large-scale convective flows, but recent observations show no such flow exists.
% second paragraph should introduce entropy rain, e.g.: In place of large-scale flows, some have suggested the solar luminosity may be transported by small ``droplets'' of rapidly falling, low entropy fluid
This ``entropy rain'' would propagate as dense vortex rings, analogous to rising buoyant ``thermals'' in the Earth's atmosphere.
In this work, we develop an analytical theory describing the evolution of dense, negatively buoyant thermals.
We verify the theory with 2D and 3D simulations of laminar, axisymmetric thermals in highly stratified atmospheres.
We discuss what the evolution of these thermals implies for the entropy rain hypothesis. % should replace this with a summary of the implications
\end{abstract}

@@

\cite{spruit1997} hypothesized that convective motions in the Sun may be driven entirely by cooling at the solar surface, and \citet{brandenburg2016} expanded upon this ``entropy rain'' hypothesis. 
Brandenburg's work includes a careful expansion of mixing length theory to incorporate flux contributions from nonlocal convective motions, and handles this theory in a horizontally-averaged sense.
He includes some discussion of possible flow morphologies which could be manifestations of this entropy rain, and even includes some brief simulations of propagating Hill vortices. 
However, these simulations and discussions did not include include a fundamental piece of entropy rain: its entropic signature deviates from the background atmosphere.
Entropy rain is dense, and the effects of antibuoyant forces will modify its dynamics.

@@

\subsection{Parameterized description of thermal evolution}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{constants.png}
    \caption{
	Evolution traces are shown for  three quantities which we assume to be constant in our parameterized thermal theory. 
	The circulation (a) and buoyancy (b), divided by their parameterized value for each case (see Table \ref{table:parameters}), are plotted vs. depth. 
	We see that with increasing stratification, there is marginally more detrainment of both of these quantities, but to first order they are constant over the	thermal's transit time. 
	In (c) the volumetric constant $f$ is shown; after significant noise during the development of the vortex ring, this quantity settles toward a value of O(7) throughout the thermal's transit.
    \label{fig:constants} }
\end{figure}

@@

We first note, as in Fig.~\ref{fig:constants}a, that the buoyancy in our thermals is not necessarily perfectly constant in time, and that effects like detrainment result in a minor reduction of the buoyant signature of our thermals. 
We thus express the buoyancy as 
\begin{equation}
B \approx \chi B_0,
\end{equation}
where $\chi$ is a constant of O(1) which represents this detrainment. 
We then integrate Eqn.~\ref{eqn:change_in_impulse} in time,
\begin{equation*}
I_z = \chi B_0 t + I_0.
\end{equation*}

@@

The momentum can likewise be simply integrated,
\begin{equation*}
M_z = \beta\chi B_0 t + M_0.
\end{equation*}
For the proper choice of vertical velocity, $w_{\text{th}}$, and volume, $\mathcal{V}$, the momentum is also precisely 
\begin{equation*}
M_z = \rho \mathcal{V} w_{\text{th}}.
\end{equation*}
We approximate the volume as $\mathcal{V} = f r^3$, where $f$ is a parameter which we take to be constant. 
This assumption is perhaps not as perfect as the assumptions of constant $B_0$ or $\Gamma_0$ (see Fig.~\ref{fig:constants}c), but it is nevertheless not a bad assumption. 
Combining our approximate expressions, and using our theoretical description of $r$ (Eqn.~\ref{eqn:r_theory}), we retrieve
\begin{equation}
\rho^{-1/2} w_{\text{th}} = \left(\frac{(\pi \Gamma)^{3/2}}{f}\right)\frac{\beta\chi B_0 t + M_0}{(\chi B_0 t + I_0)^{3/2}}.
\end{equation}
Defining the thermal velocity $w_{\text{th}} \equiv dz_{\text{th}}/dt$, and making the assumption that the vortex ring is at the virtual origin at $t = 0$ so that $I_0 = M_0 = 0$, an integrable expression for the evolution of thermal position over time can be retrieved,
\begin{equation}
\frac{dz_{\text{th}}}{\rho(z_{\text{th}})^{1/2}} =
\left(\frac{\beta(\pi\Gamma)^{3/2}}{f(\chi B_0)^{1/2}}\right)\frac{dt}{t^{1/2}}
\label{eqn:dz_theory}
\end{equation}
If the atmospheric stratification in which the thermal is falling is known, this result can be integrated with the known profile of $\rho(z_{\text{th}})$ in order to find the position of the thermal as a function of time. 
We leave this result general for now, and will integrate it using the polytropic stratification used in our simulations in the next section.

@@

\subsection{Numerics}
We evolve our simulations forward in time using the  Dedalus\footnote{\url{http://dedalus-project.org/}} pseudospectral framework \citep{burns&all2016} [CITE THE METHODS PAPER]. 
For our 2D simulations, we use using an implicit-explicit (IMEX), third-order, four-stage Runge-Kutta timestepping scheme RK443 \citep{ascher&all1997}, and for our 3D simulations we use the second-order semi-implicit backward differentiation formulation SBDF2 \citep{wang&ruuth2008}.

Our 3D simulations are decomposed on Fourier bases in the horizontal directions ($x, y \in [-L_r, L_r]$) and Chebyshev bases vertically ($z \in [0, L_z]$) with impenetrable, stress free, fixed-temperature boundary conditions at the upper and lower boundaries ($T_1 = w = \partial_z u = \partial_z v = 0$ at $z = [0, L_z]$).
Our 2D simulations are decomposed on a Fourier $(z \in [0, L_z])$ and Chebyshev $(r \in [0, L_r])$ domain, with boundary conditions of $\partial_r S_1 = w = (\grad\times\bm{u})_\phi = 0$ at $r = L_r$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% RESULTS SECTION
\section{Results}
\label{sec:results}
The measured values of each of the parameters described in section \ref{sec:theory} for each of our simulations  \ref{sec:theory} for each of our simulations  \ref{sec:theory} for each of our simulations are presented in table \ref{table:parameters}.
As anticipated, $\beta$ is O(0.5) and $\chi$ is O(1), with both decreasing slightly in value with increasing stratification, which is perhaps unsurprising in light of Fig.~\ref{fig:constants}.
In all cases, the buoyancy $B_0$ is similar to the integrated buoyancy in the initial conditions, with some losses due to detrainment in the spin-up.
In our nondimensional units, we also find that the circulation is O(-2) for each of our cases, and decreases with increasing stratification (and decreasing vortex size).

\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{results_panels.png}
    \caption{
	The measured depths of thermals as a function of time for all 2D Anelastic and 3D Fully compressible simulations conducted in this work are shown in (a).
	The corresponding thermal velocities as a function of depth are shown in (b)
	Overplotted in a thin solid line for each case are our theoretical predictions from section \ref{sec:theory} given the parameters in table \ref{table:parameters}.
    \label{fig:results_panels} }
\end{figure}

@@

In Fig. \ref{fig:results_panels}a, we show the measured depth $d_{\text{th}} = L_z - z_{\text{th}}$ of the thermal as a function of time for low and high stratification. 
At very low stratification (e.g., $n_\rho = 0.1$), the thermal is small compared to the local density scale height at all depths, and it evolves roughly according to the Boussinesq prediction of $d \propto \sqrt{t}$. 
As the stratification increases, the thermal begins to transit the domain more quickly and approaches the limit of $d \propto t^2$ predicted in the highly stratified limit of Eqn.~\ref{eqn:theory_z}. 
The theoretical fits for depth from the prediction of Eqn.~\ref{eqn:theory_z} are plotted over the measured data and show remarkable agreement.

@@

For details on how measurements of thermal properties are conducted in our simulations, we refer the reader to appendix \ref{appendix:measurements}.

@@

Integral quantities, such as the circulation, $\Gamma$, the buoyancy, $B$, and the volume, $\mathcal{V}$ require us to first determine what fraction of the domain constitutes the thermal.
To do so, we use the thermal tracking algorithm described in appendix \ref{appendix:tracking} to determine the radial contour that outlines the thermal as a function of height, $\mathcal{C}$.
We then use this contour to find our integral quantities,
\begin{equation}
\Gamma = \int_0^{\mathcal{C}}\int_0^{L_z} (\grad\times\bm{u})_\phi dz dr, \qquad
B      = \int_0^{\mathcal{C}}\int_0^{L_z} \rho S_1 dz dr, \qquad
\mathcal{V} = \int_0^{\mathcal{C}}\int_0^{L_z} 1\,dz dr.
\end{equation}

@@

\bibliography{../tex/biblio.bib}

>>>>

\begin{abstract}
Historically, large-scale convective flows called ``giant cells'' were thought to transport the Sun's luminosity from deep in the solar convection zone, but recent observations show no such flow exists.
In place of large-scale flows, some authors have suggested the solar luminosity may instead be transported by small ``droplets'' of rapidly falling, low entropy fluid.
This ``entropy rain'' would propagate as dense vortex rings, analogous to rising buoyant ``thermals'' in the Earth's atmosphere.
In this work, we develop an analytical theory describing the evolution of dense, negatively buoyant thermals.
We verify the theory with 2D and 3D simulations of laminar, axisymmetric thermals in highly stratified atmospheres.
Our results suggest that entropy rain driven at the surface of the Sun could fall into two categories: a \emph{stalling} regime in which the droplets slow down and expand, and a \emph{falling} regime in which the droplets accelerate and shrink as they propagate downwards.
We briefly discuss difficulties for the entropy rain hypothesis in both of these regimes.
\end{abstract}

@@

\cite{spruit1997} hypothesized that convective motions in the Sun may be driven entirely by cooling at the solar surface, and \citet{brandenburg2016} expanded upon this ``entropy rain'' hypothesis. 
Brandenburg's work includes a careful expansion of mixing length theory to incorporate flux contributions from nonlocal convective motions, and handles this theory in a horizontally-averaged sense.
He includes some discussion of possible flow morphologies which could be manifestations of this entropy rain, and even includes some brief simulations of propagating Hill vortices. 
However, these simulations and discussions did not include include a fundamental piece of entropy rain: its entropic signature deviates from the background atmosphere.
Entropy rain is dense, and the effects of buoyancy forces will modify its dynamics.

@@

\subsection{Parameterized description of thermal evolution}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{constants.png}
    \caption{
	Evolution traces are shown for  three quantities which we assume to be constant in our parameterized thermal theory. 
	The circulation (a) and buoyancy (b), divided by their parameterized value for each case (see Table \ref{table:parameters}), are plotted vs. depth. 
	We see that with increasing stratification, there is marginally more detrainment of both of these quantities, but to first order they are constant over the	thermal's transit time. 
	In (c) the volumetric constant $f$ is shown; after significant noise during the development of the vortex ring, this quantity settles toward a value of $\mathcal{O}(7)$ throughout the thermal's transit.
    \label{fig:constants} }
\end{figure}

@@

We first note, as in Fig.~\ref{fig:constants}a, that the buoyancy in our thermals is not necessarily perfectly constant in time, and that effects like detrainment result in a minor reduction of the buoyant signature of our thermals. 
We thus express the buoyancy as 
\begin{equation}
B \approx \chi B_0,
\end{equation}
where $\chi$ is a constant of $\mathcal{O}$(1) which represents this detrainment. 
We then integrate Eqn.~\ref{eqn:change_in_impulse} in time,
\begin{equation*}
I_z = \chi B_0 t + I_0.
\end{equation*}

@@

The momentum can likewise be simply integrated,
\begin{equation*}
M_z = \beta\chi B_0 t + M_0.
\end{equation*}
For the proper choice of vertical velocity, $w_{\text{th}}$, and volume, $\mathcal{V}$, the momentum is also precisely 
\begin{equation*}
M_z = \rho \mathcal{V} w_{\text{th}}.
\end{equation*}
We approximate the volume as $\mathcal{V} = f r^3$, where $f$ is a parameter which we take to be constant. 
This assumption is perhaps not as perfect as the assumptions of constant $B_0$ or $\Gamma_0$ (see Fig.~\ref{fig:constants}c), but it is nevertheless not a bad assumption. 
Combining our approximate expressions, and using our theoretical description of $r$ (Eqn.~\ref{eqn:r_theory}), we retrieve
\begin{equation}
\rho^{-1/2} w_{\text{th}} = \left(\frac{(\pi \Gamma)^{3/2}}{f}\right)\frac{\beta\chi B_0 t + M_0}{(\chi B_0 t + I_0)^{3/2}}.
\end{equation}
Defining the thermal velocity $w_{\text{th}} \equiv dz_{\text{th}}/dt$, and making the assumption that the vortex ring is at the virtual origin at $t = 0$ so that $I_0 = M_0 = 0$, an integrable expression for the evolution of thermal position over time can be retrieved,
\begin{equation}
\frac{dz_{\text{th}}}{\rho(z_{\text{th}})^{1/2}} =
\left(\frac{\beta(\pi\Gamma)^{3/2}}{f(\chi B_0)^{1/2}}\right)\frac{dt}{t^{1/2}}
\label{eqn:dz_theory}
\end{equation}
If the atmospheric stratification in which the thermal is falling is known, this result can be integrated with the known profile of $\rho(z_{\text{th}})$ in order to find the position of the thermal as a function of time.

To summarize, we have modeled the evolved thermal as a \emph{thin-core, axisymmetric} vortex ring which is evolving in a plane-parallel, adiabatic atmosphere, and have assumed that the motions are low Mach number such that a linearized equation of state describes the thermal well.
We have assumed that viscous and conductive effects are small and can be neglected, such that the total density-weighted specific entropy in the thermal is a conserved quantity.
We assume that vortex ring's circulation is constant, that the thermal's volume is spheroidal, and that the effects of pressure and detrainment on impulse and momentum can be parameterized as constants.
Under these assumptions, we arrived at Eqn.~\ref{eqn:dz_theory}, which we leave in its general form here, and which we will integrate using the polytropic stratification used in our simulations in the next section.

@@

\subsection{Numerics}
We evolve our simulations forward in time using the  Dedalus\footnote{\url{http://dedalus-project.org/}} pseudospectral framework \citep{burns&all2016, burns&all2019}. 
For our 2D simulations, we use using an implicit-explicit (IMEX), third-order, four-stage Runge-Kutta timestepping scheme RK443 \citep{ascher&all1997}, and for our 3D simulations we use the second-order semi-implicit backward differentiation formulation SBDF2 \citep{wang&ruuth2008}.

Our 3D simulations are decomposed on Fourier bases in the horizontal directions ($x, y \in [-L_r, L_r]$) and Chebyshev bases vertically ($z \in [0, L_z]$) with impenetrable, stress free, fixed-temperature boundary conditions at the upper and lower boundaries ($T_1 = w = \partial_z u = \partial_z v = 0$ at $z = [0, L_z]$).
Our 2D simulations are decomposed on a Fourier $(z \in [0, L_z])$ and Chebyshev $(r \in [0, L_r])$ domain, with boundary conditions of $\partial_r S_1 = w = (\grad\times\bm{u})_\phi = 0$ at $r = L_r$ and $u = \partial_r(w) = \partial_r(S_1) = 0$ at $r = 0$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% RESULTS SECTION
\section{Results}
\label{sec:results}
\begin{figure}[t!]
    \includegraphics[width=\columnwidth]{results_panels.png}
    \caption{
	The measured depths of thermals as a function of time for all 2D Anelastic and 3D Fully compressible simulations conducted in this work are shown in (a).
	The corresponding thermal velocities as a function of depth are shown in (b)
	Overplotted in a thin solid line for each case are our theoretical predictions from section \ref{sec:theory} given the parameters in table \ref{table:parameters}.
    \label{fig:results_panels} }
\end{figure}

@@

In Fig. \ref{fig:results_panels}a, we show the measured depth $d_{\text{th}} = L_z - z_{\text{th}}$ of the thermal as a function of time for low and high stratification. 
At very low stratification (e.g., $n_\rho = 0.1$), the thermal is small compared to the local density scale height at all depths, and it evolves roughly according to the Boussinesq prediction of $d \propto \sqrt{t}$. 
As the stratification increases, the thermal begins to transit the domain more quickly and approaches the limit of $d \propto t^2$ predicted in the highly stratified limit of Eqn.~\ref{eqn:theory_z}. 
The theoretical fits for depth from the prediction of Eqn.~\ref{eqn:theory_z} are plotted over the measured data.
Unsurprisingly, the theoretical fits are poor for early times when the thermal is spinning up from its initial spherical state into the vortex ring state.
Once the thermal is spun up into a vortex ring, the theory shows remarkable agreement with the measured data.

@@

The measured values of each of the parameters described in section \ref{sec:theory} for each of our simulations  \ref{sec:theory} for each of our simulations  \ref{sec:theory} for each of our simulations are presented in table \ref{table:parameters}.
As anticipated, $\beta$ is $\mathcal{O}$(0.5) and $\chi$ is $\mathcal{0}$(1), with both decreasing slightly in value with increasing stratification, which is perhaps unsurprising in light of Fig.~\ref{fig:constants}.
In all cases, the buoyancy $B_0$ is similar to the integrated buoyancy in the initial conditions, with some losses due to detrainment in the spin-up.
In our nondimensional units, we also find that the circulation is $\mathcal{0}$(-2) for each of our cases, and decreases with increasing stratification (and decreasing vortex size).

Throughout this section, the thermal's depth and radius refer respectively to the atmospheric depth and the radius from the axis of symmetry at which the vortex ring's entropy minimum is located.
The velocity is simply the derivative with respect to time of that measured depth.
For specifics on how these measurements are conducted in our simulations, we refer the reader to appendix \ref{appendix:measurements}.

@@

Integral quantities, such as the circulation, $\Gamma$, the buoyancy, $B$, and the volume, $\mathcal{V}$ require us to first determine what fraction of the domain constitutes the thermal.
To do so, we use the thermal tracking algorithm described in appendix \ref{appendix:tracking} to determine the radial contour that outlines the thermal as a function of height, $\mathcal{C}$.
We then use this contour to find our integral quantities,
\begin{equation}
\Gamma = \int_0^{\mathcal{C}}\int_0^{L_z} (\grad\times\bm{u})_\phi dz dr, \qquad
B      = 2\pi \int_0^{\mathcal{C}}\int_0^{L_z} \rho S_1 r dz dr, \qquad
\mathcal{V} = 2\pi \int_0^{\mathcal{C}}\int_0^{L_z} 1\,r dz dr.
\end{equation}

@@

\bibliography{biblio.bib}
